---
title: 'Chapter 4: Subsetting'
author: 'Charlotte Woolley'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *4.2 Selecting multiple elements*

## *4.2.1 Atomic vectors*

### 1. Positive integers return elements at the specified positions:

```{r}

(x <- c(2.1, 4.2, 3.3, 5.4))

x[c(3, 1)]

x[order(x)]

order(x)

# Duplicate indices will duplicate value
x[c(1, 1)]

# Real numbers are silently truncated to integers
x[c(2.1, 2.9)]

```


### 2. Negative integers exclude elements at the specified positions:

```{r, error=TRUE}
x[-c(3, 1)]

#Can't mix positives and negatives
x[c(-1, 2)]

# Exclude the first element
x[c(-1, 0)]

x[c(-1)]
```

### 3. Logical vectors select elements where the corresponding logical value is TRUE:

```{r}
x[c(TRUE, TRUE, FALSE, FALSE)]

x[x > 3]

#Recycling rules
x[c(TRUE, FALSE)]

x[c(TRUE, FALSE, TRUE, FALSE)]

x[c(TRUE, FALSE, TRUE)]

x[c(TRUE, TRUE, NA, FALSE)]

```

### 4. Nothing returns the original vector:

```{r}
x[]
```
### 5. Zero returns a zero-length vector:

```{r}
x[0]
```

### 6. Character vectors to return elements with matching names:
```{r}
(y <- setNames(x, letters[1:4]))

y[c("d", "c", "a")]

#You can repeat indices
y[c("a", "a", "a")]

#Need to use the exact characters
y["b"]

(z <- c(abc = 1, def = 2))
z[c("a", "d")]

#facors use the underlying integer vector not the character levels
y[factor("b")]
y[factor("d")]

```

## *4.2.2 Lists*

See 4.2.1 and 4.2.3

## *4.2.3 Matrices and arrays*

### Matrices:

```{r}

(a <- matrix(1:9, nrow = 3))

colnames(a) <- c("A", "B", "C")
a

# Select rows
a[1:2, ]

# Select columns
a[ ,1:2]

# Select logically
a[c(TRUE, FALSE, TRUE), c("B", "A")]

#Exclude
a[0, -2]


```
### Arrays:

```{r}
(vals <- outer(1:5, 1:5, FUN = "paste", sep = ","))

vals[c(4, 15)]

vals[c(1, 2)]

vals[c(-25)]

```

## *4.2.4 Data frames and tibbles*

```{r}
(df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3]))

#Columns come first
df[1]
df[1, 1]

df[df$x == 2, ]
df[c(1, 3), ]

# Select columns like a list
df[c("x", "z")]

# Select columns ike a matrix
df[, c("x", "z")]

# Different when one element
df["x"]
str(df["x"])

df[, "x"]
str(df[, "x"])

# Tibbles are special
(df2 <- tibble::tibble(x = 1:3, y = 3:1, z = letters[1:3]))
df

df2["x"]
df2[, "x"]

```

## *4.2.5 Preserving dimensionality*

```{r}
# Matrices and arrays
(a <- matrix(1:4, nrow = 2))
str(a[1, ])
str(a[1, , drop = FALSE])

# Dataframes default to drop = TRUE
(df <- data.frame(a = 1:2, b = 1:2))
str(df[, "a"])
str(df[, "a", drop = FALSE])

# Tibbles default to drop = FALSE
(df2 <- tibble::tibble(a = 1:2, b = 1:2))
str(df2[, "a"])
str(df2[, "a", drop = TRUE])

# Fators use drop differently - refers to levels
(z <- factor(c("a", "b")))
z[1]
z[1, drop = TRUE]

```

## *4.2.6 Exercises*

1. Fix each of the following common data frame subsetting errors:
```{r, error=TRUE}
mtcars

mtcars[mtcars$cyl = 4, ]
# Ans
mtcars[mtcars$cyl == 4, ]

mtcars[-1:4, ]
# Ans
mtcars[-c(1:4), ]

mtcars[mtcars$cyl <= 5]
# Ans
mtcars[mtcars$cyl <= 5, ]

mtcars[mtcars$cyl == 4 | 6, ] # returns cyls other than 4 or 6
# Ans
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]

```

2. Why does the following code yield five missing values? 
(Hint: why is it different from x[NA_real_]?)
```{r}
x <- 1:5
x[NA]
x[NA_real_]
class(NA)
class(NA_real_)

# Ans
# NA is a logical vector, so repeated/recycled to the length of x, whereas
# NA_real_ is numeric so only returns one value
```

3. What does upper.tri() return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?
```{r}
(x <- outer(1:5, 1:5, FUN = "*"))

x[upper.tri(x)]

# Ans
# upper.tri() returns entries TRUE in the upper triangle of a given matrix.
# Subsetting a matrix with it is vector style and returns a vector
# Don't need extra subsetting rules because it is suitable for matrixes
```

4. Why does mtcars[1:20] return an error? How does it differ from the similar mtcars[1:20, ]?
```{r, error = TRUE}
mtcars[1:20]
mtcars[1:20, ]
length(mtcars)

#Ans 
# mtcars[1:20] is subsetting the first 20 columns of mtcars, when there are only 11,
# so an error is returned, while mtcars[1:20, ] is asking for the first 20 rows
```

5. Implement your own function that extracts the diagonal entries from a matrix 
(it should behave like diag(x) where x is a matrix).
```{r}
# Ans
(x <- matrix(1:16, nrow = 4))
diag(x)

my_diag <- function(m){
        res <- seq(1, length(m), by = (nrow(m)+1))
        lim <- min(nrow(m),ncol(m))
        res[1:lim]
}

diag(x)
my_diag(x)

(y <- matrix(1:30, nrow = 3))
diag(y)
my_diag(y)

(z <- matrix(1:30, nrow = 10))
diag(z)
my_diag(z)

(na <- matrix(NA,nrow=3,ncol=3))
diag(na)
my_diag(na)


```

6. What does df[is.na(df)] <- 0 do? How does it work?
```{r}
(df <- data.frame(a = 1:2, b = 1:2, c = NA))
df[is.na(df)] <- 0
df

# Ans
# df[is.na(df)] <- 0 takes a subset of df where is.na(df) == TRUE and reassigns
# those values to 0

```

# *4.3 Selecting a single element*

## *4.3.1 [[*

```{r}
(x <- list(1:3, "a", 4:6))

#extract the "whole carriage"
x[1]

#extract the "carriage contents"
x[[1]]

x[1:2]
x[[c(1, 2)]]
x[[1]][[2]]
#Also
x[[1:2]]

```

## *4.3.2 $*

```{r}

# $ operator does left-right partial matching
var <- "cyl"
mtcars$var
mtcars[[var]]

x[[1]]
x$"[[1]]"

#partial matching avoided by resetting global options or using tibbles
(df <- data.frame(angus = 1:3, effie = 3:1))
df$a
df[["a"]]

(df2 <- tibble::tibble(angus = 1:3, effie = 3:1))
df2$a

#options(warnPartialMatchDollar = TRUE)

```


## *4.3.3 Missing/out of bounds indices*
```{r, error=TRUE}
x <- list(
  a = list(1, 2, 3),
  b = list(3, 4, 5)
)

purrr::pluck(x, "c", 1)
purrr::chuck(x, "c", 1)

```

## *4.3.4 @ and slot()*

See chapter 15

## *4.3.5 Exercises*

1. Brainstorm as many ways as possible to extract the third value from the cyl variable in the mtcars dataset
```{r}
#Ans
mtcars

mtcars[3,2]
mtcars[[2]][[3]]
mtcars[[c(2,3)]]
mtcars[[2:3]]
mtcars$cyl[3]
mtcars$cyl[[3]]
```
2. Given a linear model, e.g., mod <- lm(mpg ~ wt, data = mtcars), extract the residual degrees of freedom. Then extract the R squared from the model summary (summary(mod))

```{r}
mod <- lm(mpg ~ wt, data = mtcars)

#Ans
mod$df.residual
summary(mod)$r.squared

```

# *4.4 Subsetting and assignment*


```{r}
(x <- 1:5)
x[c(1, 2)] <- c(101, 102)
x

(x <- 1:5)
x[c(1, 2)] <- c(101, 102, 103)
x

(x <- 1:5)
x[c(1, 2)] <- c(101, 101, 101, 101, 101)
x

#subsetting lists with NULL are not neccesarily what you expect
(x <- list(a = 1, b = 2))
x[["b"]] <- NULL
str(x)

(x <- list(a = 1, b = 2))
x[["b"]] <- NA
str(x)

(y <- list(a = 1, b = 2))
y["b"] <- list(NULL)
str(y)


(y <- list(a = 1, b = 2))
y["b"] <- list(NA)
str(y)

#subsetting with nothing retains the structure

mtcars[] <- lapply(mtcars, as.integer)
is.data.frame(mtcars)

mtcars <- lapply(mtcars, as.integer)
is.data.frame(mtcars)
mtcars

```
# *Applications*

### *4.5.1 Lookup tables (character subsetting)*

```{r}
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = "Male", f = "Female", u = NA)
lookup[x]
unname(lookup[x])
```


### *4.5.2 Matching and merging by hand (integer subsetting)*

```{r}
grades <- c(1, 2, 2, 3, 1)

info <- data.frame(
  grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(F, F, T)
)
id <- match(grades, info$grade)
id
info[id, ]
```


### *4.5.3 Random samples/bootstraps (integer subsetting)*

```{r}
df <- data.frame(x = c(1, 2, 3, 1, 2), y = 5:1, z = letters[1:5])
df <- data.frame(x = c(1, 2, 3, 1, 2), y = 5:1, z = letters[1:5])

# Randomly reorder
df[sample(nrow(df)), ]

# Select 3 random rows
df[sample(nrow(df), 3), ]

# Select 6 bootstrap replicates
df[sample(nrow(df), 6, replace = TRUE), ]
```


### *4.5.4 Ordering (integer subsetting)*

```{r}
x <- c("b", "c", "a")
order(x)
x[order(x)]
df2 <- df[sample(nrow(df)), 3:1]
df2
df2[order(df2$x), ]
df2[, order(names(df2))]
```


### *4.5.5 Expanding aggregated counts (integer subsetting)*

```{r}
df <- data.frame(x = c(2, 4, 1), y = c(9, 11, 6), n = c(3, 5, 1))
rep(1:nrow(df), df$n)
df[rep(1:nrow(df), df$n), ]
```


### *4.5.6 Removing columns from data frames (character subsetting)*

```{r}
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df$z <- NULL
df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])
df[c("x", "y")]
df[setdiff(names(df), "z")]
```


### *4.5.7 Selecting rows based on a condition (logical subsetting)*

```{r, error=TRUE}
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]
```


### *4.5.8 Boolean algebra vs. sets (logical & integer subsetting)*

```{r}
x <- sample(10) < 4
which(x)
unwhich <- function(x, n) {
  out <- rep_len(FALSE, n)
  out[x] <- TRUE
  out
}
unwhich(which(x), 10)

(x1 <- 1:10 %% 2 == 0)
(x2 <- which(x1))
(y1 <- 1:10 %% 5 == 0)
(y2 <- which(y1))
x1 & y1
intersect(x2, y2)
x1 | y1
union(x2, y2)
x1 & !y1
setdiff(x2, y2)
xor(x1, y1)
setdiff(union(x2, y2), intersect(x2, y2))

```


### *4.5.9 Exercises*

1. How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?
```{r}
#Ans
VADeaths <- as.data.frame(VADeaths)
VADeaths[ , sample(ncol(VADeaths), ncol(VADeaths))]
VADeaths[sample(nrow(VADeaths), nrow(VADeaths)), ]
VADeaths[sample(nrow(VADeaths), nrow(VADeaths)), sample(ncol(VADeaths), ncol(VADeaths))]
```


2. How would you select a random sample of m rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?
```{r}
x <- tibble::tibble(row = 1:10, names = c("angus", "effie", "neville", "paris", "gypsy", "georgia", "indy", "ty", "lacey", "spencer"))
x

#Ans 
simple_sample <- function(df, m) {
df[sample(nrow(df), m), ]
}

simple_sample(x, 5)
simple_sample(x, 2)
simple_sample(x, 7)

contiguous_sample <- function(df, m) {
s_val <- sample(nrow(df) - m + 1, 1)
df[s_val:(s_val + m - 1), ]
}

contiguous_sample(x, 5)
contiguous_sample(x, 2)
contiguous_sample(x, 7)
```

3. How could you put the columns in a data frame in alphabetical order?
```{r}
VADeaths <- as.data.frame(VADeaths)
VADeaths

#Ans
VADeaths[order(names(VADeaths))]
```


# *4.6 Chapter Answers*

1. What is the result of subsetting a vector with positive integers, negative integers, a logical vector, or a character vector?

"Positive integers select elements at specific positions, negative integers drop elements; logical vectors keep elements at positions corresponding to TRUE; character vectors select elements with matching names.""

2. What’s the difference between [, [[, and \$ when applied to a list?

"[ selects sub-lists: it always returns a list. If you use it with a single positive integer, it returns a list of length one. [[ selects an element within a list. \$  is a convenient shorthand: x\$y is equivalent to x[["y"]]"

3. When should you use drop = FALSE?

"Use drop = FALSE if you are subsetting a matrix, array, or data frame and you want to preserve the original dimensions. You should almost always use it when subsetting inside a function."

4. If x is a matrix, what does x[] <- 0 do? How is it different from x <- 0?

"If x is a matrix, x[] <- 0 will replace every element with 0, keeping the same number of rows and columns. In contrast, x <- 0 completely replaces the matrix with the value 0."

5. How can you use a named vector to relabel categorical variables?

"A named character vector can act as a simple lookup table: c(x = 1, y = 2, z = 3)[c("y", "z", "x")]"




